Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-30 vertical-thermal-review.Rmd
improving ecology section

Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010]. 
Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
The observation that both the abundance of emitting species (Table ) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
In the meantime, our best understannding is that isoprenes are important for mid-canopy thermal regulation, but that other traits are more important in the most sun-exposed canopy leaves.

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates, measured as radial stem growth or biomass increment, generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006; @stephensonRateTreeCarbon2014]. 
NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From Anderson-Teixeira et al. (in revision). In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020. ](tree_rings/Figure_4_tree_rings.png)

>>>>

Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010]. 
Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
The observation that both the abundance of emitting species (Table 1) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
In the meantime, our best understannding is that isoprenes are important for mid-canopy thermal regulation, but that other traits are more important in the most sun-exposed canopy leaves.

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russell_residence_2014].
While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2021], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2021], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.

Vertical gradients also affect the climate sensitivity of metabolism and growth.
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. 
This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. 
This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimate2021. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020. ](tree_rings/Figure_4_tree_rings.png)
